Reconstructing the last known movements of one of Nature's giants

### 2017-06-23 manuscript/blue-whale-ms.tex
Tiny remaining edits from CT

\subsection{Agent-based whale movement model}
Movement was coded as set of probabilistic rules. 
The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
All terms were expressed as probability distributions allowing us to introduce individual variability.

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$) of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
Northerly migrations were possible only in spring, and southerly migrations in autumn. Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
(ii) Sea surface temperature ($^{\circ}$C\cite{yool2013medusa}). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters. 
(iii) Water depth (m; what reference is this collected from?). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes. 
Various movement rules were implemented to avoid simulations ending up on land, or becoming `stuck' in the relatively shallow waters of the Bay of Biscay or the Strait of Dover. 
For a full description of the model and its parameters see Supplementary Methods. 
We simulated whale movements XX times for each scenario and residency hotspot. 
For each movement simulation we extracted phytoplankton $\delta^{13}$C and $\delta^{15}$N values at each time point and location from the models described above\cite{magozzi2017using} (Somes). 
We then compared the simulated stable isotope profiles (Fig. S4) to the profile of the blue whale (Fig. 1).

\subsection{Data Availability}
Data are available on the Natural History Museum, London's data portal (https://data.nhm.ac.uk). 
R code is available from XXX.

@@

\section{Acknowledgments}
% adjectives not allowed!
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank XXX for comments on drafts.
We thank CJ Somes for providing  $\delta^{15}$N POM data.

@@

\section{Author information}
Reprints and permissions information is available at www.nature.com/reprints.
The authors declare no competing financial interests.
Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.
% One person only allowed

%\section{Tables}

@@

%\newpage
%\section{Figures}

%  \begin{figure}[!htbp]
%    \centering
%      \includegraphics[width=12cm]{XXX.pdf}
%      \caption{}
%      \label{}
%  \end{figure}

>>>>

\subsection{Agent-based whale movement model}
We simulated the movement of the whale with the likelihood, direction, and extent of movement influenced by behavioural state, sea surface temperature, water depth, and phytoplankton concentration (as a proxy for zooplankton food availability). 
At each time point and location in the simulations, we extracted phytoplankton $\delta^{13}$C values from the spatio-temporal model\cite{magozzi2017using} resulting in a range of simulated isotopic profiles for different migration patterns and foraging areas. 
Movement was coded as set of probabilistic rules. 
The parameters for these were taken from the literature on blue whale behaviour (e.g.\cite{handbook}). 
All terms were expressed as probability distributions allowing us to introduce individual variability.

In the models, the likelihood of movement, direction (north, south, east, west, northeast, northwest, southeast, or southwest) and extent (1$^{\circ}$ or 2$^{\circ}$) of movement are all influenced by the following. 
(i) Behavioural state (migrating north, migrating south, foraging, or nursing). 
Northerly migrations were possible only in spring, and southerly migrations in autumn. Foraging was possible at anytime of year, and was triggered when whales encountered high concentrations of plankton. Nursing occurred at temperatures between 18 and 25$^{\circ}$C only. 
(ii) Sea surface temperature ($^{\circ}$C\cite{yool2013medusa}). 
When migrating north, whales were more likely to move towards lower temperatures provided they were above the minimum temperature threshold (3$^{\circ}$C); whereas whales migrating south sought warmer waters. 
(iii) Water depth (m; from the General Bathymetric Chart of the Oceans (GEBCO) global bathymetry dataset). 
Whales were less likely to move into waters less than 400m deep, and increasingly unlikely to move into even shallower waters. 
(iv) Phytoplankton concentration (mmol N $m^{-3}$, for combined diatom and non-diatom communities\cite{yool2013medusa}). 
This was included as a proxy for zooplankton food availability. 
Whales are more likely to move towards (or remain within) areas of high phytoplankton density, particularly during the foraging behavioural state.

We simulated the isotopic expression expected for (a) residency in each known hotspot for blue whale sightings or historic hunting grounds in the North Atlantic (Norwegian/Barents Sea, West Ireland, South Greenland, Nova Scotia/Newfoundland and Canaries/Azores\cite{mcdonald2006biogeographic,reilly2008balaenoptera,sigurjonsson1995life}; Fig SX; (b) seasonal migration between high sub-arctic latitudes and temperate latitudes around the British Isles and (c) seasonal migration between high latitudes and subtropical latitudes. 
For a full description of the model and its parameters see Supplementary Methods. 
We simulated whale movements 30 times for each scenario and residency hotspot. 
For each movement simulation we extracted phytoplankton $\delta^{13}$C values at each time point and location from the models described above\cite{magozzi2017using}. 
We then compared the simulated stable isotope profiles (Fig. \ref{fig3}) to the profile of the blue whale (Fig. \ref{fig1}).

\subsection{Data Availability}
Data are available on the Natural History Museum, London's data portal (https://data.nhm.ac.uk). 
R code is available from GitHub (github.com/ALJackson/blue-whale-bes).

@@

\section{Acknowledgments}
% adjectives not allowed!
This work was funded by the British Ecological Society (grant: 5771/6815). 
We thank XXX for comments on drafts.
We thank C.J. Somes for providing  $\delta^{15}$N POM data.

@@

\section{Author information}
Reprints and permissions information is available at www.nature.com/reprints.
The authors declare no competing financial interests.
Correspondence and requests for materials should be addressed to natalie.cooper@nhm.ac.uk.

@@

%\section{Figures}
